Compilation with Dynamic Grammars

### 2017-04-22 papers/paper.tex
Adds paper/notes, third draft

Since each programming language has its own compiler, there is some redundancy in each implementation. 
For example, the lexer for a C++ compiler and a Java compiler would be extremely similar, since both languages use braces, semicolons, and parentheses in the same way.
Even two programming languages that don't have similar syntax have similar parts of their compiler. 
For example, a lexer seperates text based on a list of delimiter-strings, which vary with each programming language. The string seperation function, however, is the same for every lexer, and would be redundant across multiple compilers.
The same concept applies to Parsers and Generators, etc.. (Parsers read tokens and produce an ast, Generators use an AST to generate code)

If programming languages could be compiled with a framework, this redundancy would go away, and programming language researchers would be able to focus on the language itself, as opposed to its compiler.
Essentially, Glossa abstracts the parts of a compiler that don't depend on which language is being compiled, and reads in rules for each language, effectively seperating the language from how it is compiled. 
Because of this, Glossa allows simple implementations of programming languages, which means that Glossa can easily become capable of translating many different languages with ease.

@@

    \node (sourcea) [startstop]                       {Source Code A};
    \node (lexer)   [process, below of=sourcea]       {Lexer};
    \node (tokens)  [io, below of=lexer]              {Symbolic Tokens};
    \node (lrules)  [process, right of=lexer, xshift=2cm]  {Lex Rules};
    \node (lfiles)  [io, right of=lrules, xshift=2cm]  {Lex Files};
    \node (parser)  [process, below of=tokens]        {Parser};
    \node (asta)    [io, below of=parser]             {AST};
    \node (grules)  [process, right of=parser, xshift=2cm]  {Grammar Rules};
    \node (gfiles)  [io, right of=grules, xshift=2cm] {Grammar Files};
    \node (trans)   [process, below of=asta]          {Transformer};
    \node (astb)    [io, below of=trans]              {Intermediate AST};
    \node (trules)  [process, right of=trans, xshift=2cm]  {Transformer Rules};
    \node (tfiles)  [io, right of=trules, xshift=2cm]  {Transformer Files};
    \node (gen)     [process, below of=astb]          {Generator};
    \node (stop)    [startstop, below of=gen]         {Source Code B};
    \node (ggrules) [process, right of=gen, xshift=2cm]  {Generator Rules};
    \node (ggfiles) [io, right of=ggrules, xshift=2cm]    {Generator Files};

@@

Transformations to the AST allow for an overall simpler compilation process, reducing the number of construction files required for an output language. 
AST transformations also solve non-isomorphisms, which is very useful if two languages are not very similar.

Essentially, transformations change language-specific features into more generalized ones. 
AST transforms can also model code optimizations.

Each element of the AST is MultiSymbol, which is a tag (string), and a string dictionary of symbols.
To transform an AST element, the tag can be changed, or elements within the string dictionary can be manipulated.

A transformer file is just a series of these transformations:
\begin{verbatim}
add    key (literal | identifier | ..) (2 | x | ...)
append key (literal | identifier | ..) (2 | x | ...)
copy   key1 key2
move   key1 key2
copy-append key1 key2
move-append key1 key2
retag newtag
\end{verbatim}
Which can be surrounded by control flow, in the same way that constructor files can:
\begin{verbatim}
branch defined identifier 
move identifier old_identifier
add name identifier f
elsebranch
add name identifier f
end
\end{verbatim}
(This AST transform renames the syntax element to f, saving the old name to the key "old\_identifier"

@@

If Glossa has adequate language descriptions, then translations are trivial. 
Since Glossa can convert one language to another, it works as a pre-processor.
Glossa has the ability for languages to inherit from and extend other languages, so, for example, C++ might inherit from C, and then implement class statements that compile into C code.

@@

Since Glossa uses dynamic language descriptions, languages can easily be created or modified by changing their descriptor files. Glossa also allows languages to inherit from existing languages. For example, Glossa's descriptions for C++ might inherit from C, and only overwrite or add features.
This functionality makes language experimentation painless, as code doesn't even need to be recompiled.

@@

This example isn't necessarily meant to be an actual programming language, but more to show how a language could easily be implemented in Glossa.

@@

Glossa shows that a multi-language source to source compiler is possible. 
Since description files are the only pre-requisite for translating a particular language, Glossa can theoretically translate any language, as long as a description is provided.
While Glossa only supports a few different languages, it could easily support more if more description files were written.
However, Glossa's interface would benefit from simplification, since writing description files is a hard task. Also, even in description files, there is some redundancy that could be eliminated.

Glossa has accomplished far more than it originally set out to: Perhaps the software will mature in the future, and eventually be used at the industry level.

>>>>

Since each programming language has its own compiler, there is some redundancy in each implementation. 
Essentially, Glossa abstracts the parts of a compiler that don't depend on which language is being compiled, and reads in rules for each language.
Because of this, Glossa allows simple implementations of programming languages, which means that Glossa can easily become capable of translating many different languages with ease.
For example, the lexer for a C++ compiler and a Java compiler would be extremely similar, since both languages use braces, semicolons, and parentheses in the same way.
Even two programming languages that don't have similar syntax have similar parts of their compiler. 
For example, a lexer seperates text based on a list of delimiter-strings, which vary with each programming language. The string seperation function, however, is the same for every lexer, and would be redundant across multiple compilers.
The same concept applies to Parsers and Generators, etc.. (Parsers read tokens and produce an AST, Generators use an AST to generate code)

If programming languages could be compiled with a framework, this redundancy would go away, and programming language researchers would be able to focus on the language itself, as opposed to its compiler.

@@

    \node (sourcea) [startstop]                       {Source Code A};
    \node (lexer)   [process, below of=sourcea]       {Lexer};
    \node (tokens)  [io, below of=lexer]              {Symbolic Tokens};
    \node (lrules)  [process, right of=lexer, xshift=2cm]  {Lex Rules};
    \node (lfiles)  [io, right of=lrules, xshift=2cm]  {Lex Files};
e   \node (parser)  [process, below of=tokens]        {Parser};
    \node (asta)    [io, below of=parser]             {AST};
    \node (grules)  [process, right of=parser, xshift=2cm]  {Grammar Rules};
    \node (gfiles)  [io, right of=grules, xshift=2cm] {Grammar Files};
    \node (trans)   [process, below of=asta]          {Transformer};
    \node (astb)    [io, below of=trans]              {Intermediate AST};
    \node (trules)  [process, right of=trans, xshift=2cm]  {Transformer Rules};
    \node (tfiles)  [io, right of=trules, xshift=2cm]  {Transformer Files};
    \node (gen)     [process, below of=astb]          {Generator};
    \node (stop)    [startstop, below of=gen]         {Source Code B};
    \node (ggrules) [process, right of=gen, xshift=2cm]  {Generator Rules};
    \node (ggfiles) [io, right of=ggrules, xshift=2cm]    {Generator Files};

@@

AST Transformations solve non-isomorphisms, making language conversions more flexible by reducing language-to-language dependencies.
Essentially, languages with a specialized feature set convert their language features to generalized ones (i.e. Haskell pattern matching might become a series of conditional checks on the arguments of a function). (Initial AST -\textgreater Universal AST)
Then, if the output language does not contain some features in the Universal Feature Set, then it represents the same concept in its native form. (Universal AST -\textgreater Specialized AST).

For example: 
Haskell's Pre-transformer might convert a parsed \textit{Data} element into a universal structure like a class.
(Haskell's special data type is not in the Universal Feature set, but a class is.)
Then, if the output is in a language that doesn't support classes, i.e. C, then C's Post-transformer would convert the class to struct-function-collections, creating the specialized AST, which would be compilable in C.

\newpage
\begin{tikzpicture}[node distance=2.5cm]

    \node (asta)   [io]                       {AST};
    \node (transa) [process, below of=asta]   {Pre-transformer};
    \node (astb)   [io, below of=transa]       {Universal AST};
    \node (transb) [process, below of=astb]   {Post-Transformer};
    \node (astc)   [io, below of=transb]       {Specialized AST};

    \draw [arrow] (asta)   -- (transa);
    \draw [arrow] (transa) -- (astb);
    \draw [arrow] (astb)   -- (transb);
    \draw [arrow] (transb) -- (astc);

    % title
    \node[align=center,font=\bfseries, yshift=2em] (title) 
	at (current bounding box.north)
	{Dual-AST transform};

\end{tikzpicture}
\newpage

Each element of the AST is MultiSymbol, which is a tag (string), and a string dictionary of symbols. Each of these symbols can be either another MultiSymbol, or an atom like a literal, operator, or identifier.
To transform an AST element, the tag can be changed, or elements within the dictionary can be manipulated.

A transformer file is just a series of these transformations:
\begin{verbatim}
createreg [name]
// Copy an element from global multisymbol table to a key in a register's multisymbol table
reg [name] transfer(-append) [reg-key] [global-key]
// Move the multisymbol from a register to the destination key in the multisymbol table. If a destination_reg is specified, this table is the table of a register. Otherwise, it is the global multisymbol table
pushback [reg-name] [destination] (optional)[destination_reg]
// Prefix for performing an operation on a register
reg [name] {command (any of below)}

// The following also work with operations on the table of the current multisymbol
// In add and append, type might be "identifier" or "literal" and value might be "x" or "2"
// I.e to add an identifier foo under the key "name", use the command (add name identifier foo)
// Add overwrites a key, append adds onto it.
add    [key] [type] [value]
append [key] [type] [value]
// Move deletes the source key, copy does not.
copy(-append) [key1] [key2]
move(-append) [key1] [key2]
// Change the tag of the register
retag newtag
// Delete a key from the multisymbol table
delete [key]
\end{verbatim}
Which can be surrounded by control flow, in the same way that constructor files can:
\begin{verbatim}
branch defined identifier 
move identifier old_identifier
add name identifier f
elsebranch
add name identifier f
end
\end{verbatim}
(This AST transform renames the syntax element to f, saving the old name to the key "old\_identifier"

@@

Because Glossa dynamically reads in language descriptions, it can support a large number of programming languages.
When given adequate description files, Glossa is potentially as capable as a specialized compiler or interpreter.
Several examples of Glossa's translation capabilities are shown in the Appendix.

Even though translation works very well, there are still several limitations to keep in mind:
Retention of readability is not always garunteed, especially if there are many non-isomorphisms that need to be solved.
Code that uses libraries is harder to translate: 
It requires either a foreign function interface to be written, or the library to be translated alongside the software. 
Translation of the library depends on access to the library's source code, which isn't always available.

@@

Since Glossa uses dynamic language descriptions, languages can easily be created or modified by changing their descriptor files. 
This functionality encourages language experimentation.
Glossa also allows languages to inherit from existing languages. 
For example, Glossa's descriptions for C++ might inherit from C, and only overwrite or add features.

Language design in Glossa is not affected in the same way that Translation is.
The only possible limitation is the descriptor file interface.
The success of an implemented language depends on how well it has been described.
If description is straightforward, then language creation will be as well.

@@

This is only an example of a DSL (Domain Specific Language), but a full language could easily be implemented using the exact same concepts.

@@

Glossa shows that a multi-language source to source compiler is possible. 
Since description files are the only pre-requisite for translating a particular language, Glossa can theoretically translate any language, as long as a description is provided.
While Glossa only supports a few different languages, it could easily support more if more description files were written.
However, Glossa's interface would benefit from simplification, since writing description files is a hard task. 
Also, even in description files, there is some redundancy that could be eliminated.
As the description file interface improves, Glossa will come closer to being an idealistic multi-language source to source compiler.

While Glossa is useful at doing some automated translations, it is likely more useful for implementing new programming languages.
Because of the generalizations (discarding of hard-coded rules) that Glossa makes, changing language definitions is trivial compared to the potential difficulty of modifying the source of a compiler.
With Glossa, a new language can be implemented in a minimal amount of time, and then modified easily.

Glossa has accomplished far more than it originally set out to: Perhaps as its interface improves, Glossa can be used as groundwork for an industrial-level compiler framework.
